---
title: "Flight delay"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flight delay}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(BonusLab)
library(nycflights13)
library(dplyr)
library(caret)
library(knitr)
```

## Data
### Loading the flight and weather dataset
Both the weather and flights dataset is from the nycflights13 package.
```{r}
data(weather)
data(flights)
```

### Mergeing the data sets
The two data sets are merged by time period and origin. Then the departure delay, origin, temperature, wind speed, precipiation, air pressure and visablity variables are selected. Two interaction varaibles, windspeed-precipiation and windspeed-visablity are also created. 
```{r}
# data with weather and flights (delays)
dww <- left_join(flights, weather, by=c("time_hour"="time_hour", "origin"="origin")) %>%
 select(dep_delay, origin, temp, wind_speed, precip, pressure, visib) %>%
 stats::na.omit() %>% mutate(wind_pre = wind_speed*precip,
                             wind_vis = wind_speed*visib)
```

### Data split
The data set are then split up into training (80%), validation (15%) and test (5%).
```{r}
# row in data: 100%
rows_in_data <- 1:nrow(dww)

# test or validation: 100%*0.2 = 20%
test_val_Index <- caret::createDataPartition(rows_in_data, p = .2, 
                                  list = FALSE, 
                                  times = 1)
# Train: 100%-20% = 80%
train_index <- rows_in_data[-test_val_Index]

# validation: 0.75*20% = 15%
val_index <- caret::createDataPartition(test_val_Index, p = .75, 
                                  list = FALSE, 
                                  times = 1)

# test: 0.25*20% = 5%
test_index <- test_val_Index[-val_index]
```


```{r, echo=FALSE}
df_data<-data.frame(val=(length(val_index)/length(rows_in_data))*100,
           train=(length(train_index)/length(rows_in_data))*100,
           test=(length(test_index)/length(rows_in_data))*100)

colnames(df_data)<-c("Validation", "Training", "Test")
rownames(df_data)<-c("")
kable(df_data)
```
The table above shows the acctual size (%) of the three datasets.

```{r}
# Datasets
train_data <- as.data.frame(dww[train_index,])
val_data <- as.data.frame(dww[val_index,])
test_data <- as.data.frame(dww[test_index,])
```

## Training the models
To find an optimum value of \(\lambda\), four models where trained with different values of \(\lambda\). The models will be compared by validation RMSE.

```{r}
# Function to calculate RMSE
RMSE_flight_delay <- function(model, validation=TRUE){
  if(validation==FALSE){
    data <- test_data
  } else if (validation==TRUE){
    data <- val_data
  }
  
  actual <- data$dep_delay
  pred <- model$predict(data)
  
  return(sqrt(sum((pred-actual)^2/length(actual))))
}
```


```{r}
# Lambda = 1
model1 <- ridgereg(data=train_data,
                   formula=dep_delay~origin+temp+wind_speed+precip+pressure+visib+wind_pre+wind_vis,
                   lambda=1)
RMSE1 <- RMSE_flight_delay(model1, validation=TRUE)

# Lambda = 5
model2 <- ridgereg(data=train_data,
                   formula=dep_delay~origin+temp+wind_speed+precip+pressure+visib+wind_pre+wind_vis,
                   lambda=5)
RMSE2 <- RMSE_flight_delay(model2, validation=TRUE)

# Lambda = 10
model3 <- ridgereg(data=train_data,
                   formula=dep_delay~origin+temp+wind_speed+precip+pressure+visib+wind_pre+wind_vis,
                   lambda=10)
RMSE3 <- RMSE_flight_delay(model3, validation=TRUE)

# Lambda = 25
model4 <- ridgereg(data=train_data,
                   formula=dep_delay~origin+temp+wind_speed+precip+pressure+visib+wind_pre+wind_vis,
                   lambda=25)
RMSE4 <- RMSE_flight_delay(model4, validation=TRUE)
```


```{r, echo=FALSE}
RMSE_table <- data.frame(model=c(1:4),
                         lambda=c(1,5,10,25),
           `Validation RMSE`=c(RMSE1, RMSE2, RMSE3, RMSE4))
colnames(RMSE_table)[2:3]<-c("lambda", "Validation RMSE")
kable(RMSE_table)
```

The table above shows the RMSE for the four different models. The best model (lowest RMSE) is model number 3 with \(\lambda=10\)


## Test RMSE
```{r}
RMSE_flight_delay(model3, validation=FALSE)
```
RMSE for the validation data is about X.
